Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-04-06 WSSSPE2_report/WSSSPE2_review.tex
adding stuff from Nancy

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Papers}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
The papers that were discussed in the Software Development Experiences group are:
\begin{itemize}
\item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric
Hutton, Gregory Tucker and Erkan Istanbulluoglu. Landlab: Sustainable Software
Development in Practice ~\cite{wssspe2_adams}

@@

\item Michael R. Crusoe and C.Titus Brown. Channeling community contributions to
scientific software: a hackathon experience~\cite{wssspe2_crusoe}

>>>>

\katznote{need some short general intro to Software Development Experiences here}

@@

Because of the large number of submitted papers categorized as Software
Development Experiences, papers were split into two subgroups prior to the
workshop.

\katznote{need something about subgroup A too}

Subgroup B discussed four papers and had six
participants. Because of the nature of the papers, training emerged as a
common theme, however the conversation was wide-ranging, including
incentives, reproducibility and funding to promote sustainability.

\subsubsection{Subgroup A Discussion}

\subsubsection{Subgroup B Discussion}
The statement ``applied computer science is being attempted in academia
without any formal training'' kicked off our discussion. The group brought
expertise in several different training models, from a 2-day Software
Carpentry workshop, to a 3-week Open Science for Synthesis (OSS) program
to semester-long programs. We discussed the pros and cons of different
training approaches, also touching on informal learning, for example where
people learning the necessary skills because they had cross-disciplinary
people (``boundary scientists'') in their work environments of whom they
could ask questions.

Some papers explored gaps in training of early career scientists. Industry
participants in our group confirm this. We asked ourselves, ``Are
traditional courses failing?''. We think yes. Changes to undergraduate
curricula requirements are difficult. But as programming models become
more complex, we have to raise the skill levels. Skills must be improved
not only in traditional programming - learning languages and algorithms -
but around professional software development. We want to get to a point
where ``Everyone has a new minimum now - everyone knows git''. Developers
also need training on licensing choices. Just because a code is on git
doesn't mean it's open.

"How do you get people to look for the training they need?'' wondered one
participant. People seek out opportunities like Software Carpentry to
augment skills. While some instruction can be done in an institutional
curricula, independent groups (non-profits, institutes) have more
flexibility. Some asked whether Silicon Valley be interested in funding
training so that people are better prepared to enter the workforce? Some
felt that companies were reluctant to deliver training because employees
then leave the company. While others felt that pushing all training to
industry could lead to good technical people ``getting on the Google bus''
and leaving the sciences.

"How will we know when people are trained effectively in these new
skills?" We discussed certification. It can be hard to build the
recognition of Java, database certifications amongst all technologies. OSS
offers badges to those completing training, however we need to demonstrate
proficiency, not just completion. Google Summer of Code is a big CV
augmentation for participants, can we duplicate something like that?

The group also discussed how to fund training. On participant observed
that NYU runs a 6-week data science training, companies grab graduates but
pay for those they hire. OSS also used the NSF Software Institutes program
as a vehicle to fund training activities. Software Carpentry uses a
collaborative teaching approach where  people publish open teaching
materials and receive credit for their use.

We then discussed career paths for those supporting sustainable software.
While tenure track is not the only option for students, the challenges for
those who remain in academia can be large. Research scientists are
entirely dependent on soft money, which can be unpredictable. Postdocs and
those who do pursue tenure track positions need to publish and see no
rewards in software development. These challenges were all identified at
WSSSPE1. What actions would we recommend to improve things? Altmetrics and
download statistics may slowly change the system and improving a
developer's ability receive credit for time invested in software
development. NSF recognition of data products (including software) as well
as publications is also helping.

We asked ourselves, ``Where are there examples of things that are changing
because of this and how can we build momentum?''. One example demonstrates
change over time. In 2007, nanoHUB listed the academic reward structure as
a problem in an EDUCAUSE report where the authors note, ``In the future,
nanoHUB researchers are hoping to change the research culture. While they
recognize that young faculty members are unlikely to get tenure based on
their nanoHUB contributions, they hope to encourage faculty to think
beyond their own research needs to consider publishing tutorials and other
Web content in their field on the nanoHUB site''. Fast forward seven years
later where, in a 2014 iSGTW article, quote nanoHUB PI Gerhard Klimeck, ``A
former student of mine published eight tools on nanoHUB, serving over
6,000 people with his tools. He then joined a university as a professor
and introduced nanoHUB. Use of the gateway from that university
skyrocketed; he used nanoHUB in existing classes, created new classes, and
infused it in his research. Ultimately, the professor's department head
associates his two-year rise to tenure with the notoriety and innovation
he gained through nanoHUB.''

The group also discussed ``attribution trees'', an idea put forward by Dan
Katz and Arfon Smith where a chain of attributions can, for example, give
credit to those developing libraries and building blocks that other pieces
of software use. The group considered where to push the use of this, which
journals? On participant noted that Dryad works with the journals. If a
paper is accepted, the supporting data must be submitted. We discussed
reproducibility as a component of the journal review process and ``active
papers'', with immediate links to the data and software.

We then discussed variations amongst scientific domain areas and wondered
"Are some communities more or less open than others?'' To some members of
the group biology seems to be open, while physics less so. Some felt with
the more recent development of bioinformatics as a field there were fewer
historical practices to undo. Physics has pre-print philosophy to
overcome. Environmental sciences may be mixed. The group felt the
biomedical area, however, was very competitive and closed. One participant
mentioned blueprints for going open source (like NWChem recently did)
where authors outline how this helps, what you do and what the next steps
are.

We then moved beyond our training discussion to address funding that
encourages sustainable software -- funding of both people and of projects
that create a true system to support sustainability. We called this
``institutionalized serendipity''. Because science is increasingly reliant
on software, one participant observed that ``software development can have
much broader impact than publishing individual research, but it is not
viewed that way''. Because of this centrality, one participant mentioned
that training ought to be called Science Carpentry rather than Software
Carpentry. We felt we were beginning to see changes in the research
community as a result of the NSF's data management plan requirements. PIs
are thinking more about data and some university libraries are offering
data repositories. We wondered if a software management plan might be
effective. "How might funding programs need to adapt to reward good
software development practices?'', we asked ourselves. We thought about
best practices such as version control, test harnesses, mailing lists, bug
tracking, community contribution, and reuse where appropriate. We thought
about measuring success through usage statistics (downloads, altmetrics).
"Should funders demand reproducibility?'' we wondered? In order for results
to be reproduced, software would need to be carefully curated.

Again, our industry participants contributed unique viewpoints. Partnering
with industry was seen one path to sustainability. The unique partnerships
Kitware engages in promote academic freedom while creating an income
stream from certain portions of the software. This type of approach to
sustainable software frees researchers from performing tasks that do not
offer the rewards their institution values''. We talked some about
successful models for industry partnerships that preserve open science.
Participants noted that there are some NSF programs that prohibit
partnerships with for profit companies (but there are others where this is
encouraged).

\subsubsection{Subgroup A Actions}

\subsubsection{Subgroup B Actions}

Subgroup B then focused on actions it could take. It discussed development
of a white paper that describes a matrixed approach to training
(multi-day, multi-week, semester). The white paper might include a survey
of existing techniques. There are many, some dating back many years, for
example the Interuniversity Consortium for Political and Social Research
(ICPSR) and various summer institutes. The white paper could Include a
call for a comprehensive assessment of these techniques. We need to think
carefully about the right venue for such a white paper, where would it
have the most impact. The subgroup believes it would need to approach editors
directly to ascertain this.

The group felt that training in techniques that promote sustainability has
a range of benefits: career paths, educated reviewers, reproducible
science and more. There is some information on how that has been
approached and assessed, but more is needed. The group felt that this
training is undervalued and that it's important to communicate the return
on investment - both individual ROI (skills that make scientists more
effective and more marketable) and funder ROI (better use of taxpayer
funds, research more likely to be reproducible because sustainable
software exists, better trained reviewers).

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Papers}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
The papers that were discussed in the Software Development Experiences Subgroup A are:

\begin{itemize}
\item Michael R. Crusoe and C.Titus Brown. Channeling community contributions to
scientific software: a hackathon experience~\cite{wssspe2_crusoe}

The papers that were discussed in the Software Development Experiences Subgroup B are:

\begin{itemize}

\item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric
Hutton, Gregory Tucker and Erkan Istanbulluoglu. Landlab: Sustainable Software
Development in Practice ~\cite{wssspe2_adams}

@@
